Finite Element Method
A deterministic numeric procedure applied to PDEs

A PDE is a model, and like every model is incorrect, i.e. misspecified:
elastic versus plastic, absence of vortices, physics of turbulence is not yet  understood, material. Naiver-Stokes thus is usually (always) misspecified.

\begin{itemize}
    \item Miss-specification (oh-hoh)
    \item Exploit the ability to make measurements / generate data.
\end{itemize}


How to take data and incorporate it into FEM, allowing it to update
the FEM solution.

* embed in to probabilistic framework and compute uncertainty propagation

Redefine FEM in a statistical framework

The premise: there is some truth, that is known partially, and not
fully understood. But we can measure it (as generated from truth).

Cox's thesis about iterative model-observation-hypothesis cycle.

PDE BVRP
$$
    \Lcal_\theta u + f = 0
    \text{ on } \Omega
    \,, \\
    \Dcal u = 0 \text{ on } \partial \Omega
    \,. $$

Diff. operators $\Lcal__\theta$ $\Dcal$ mapping Hilbert spaces into
subsets of $\Lcal^2(\Omega)$.

Randomized setup: the solution and the operators depend on Tyche's choice.
$$
    \Lcal_\theta(\omega) u(x, \omega) + f(x, \omega) = 0
    \forall x\in \Omega
    \,. $$

Gaussian measure on the function (Hilbert space): a G measure conditional
on $\theta$ (random function) and the action of $\Lcal_\theta$ on $u$
is given by a Gaussian process, i.e $f\sim \mathcal{GP}(0, K_f)$.

This pushforward induces a measure of the space of solutions:
$$
u \sim p(u\vert f)
    = \mathcal{GP}\bigl(
    -(\Lcal_\theta)^{-1} f, 
    (\Lcal_\theta)^{-\top} K_f (\Lcal_\theta)^{-1}
    \bigr)
    \,. $$
We look for weak solutions (eq. in \mathbf{Law}) via Portmanteau (?).

FEM -- $u, f$ expansion via some basis functions, mesh resolution $h$
FEM -- $\sum_i u_i B_\theta = \lale\rale$
This induces a finite-dim conditional Gaussian measure
$$
p(u_h \vert f, \theta, h)
    = \mathcal{N}(
        A_\theta^{-1}, A_\theta^{-1} G A_\theta^{-1}
    )
    \,, $$
and a non-gaussian measure marginalizing $\theta$.

The matrix $A_\theta$ is given by the bilinear form (in the weak solution),
$G$ is Gram-matrix of the $K_f$-inner-product of the basis elements.

The measure is centered around the Galerkin solution of the FEM,
and the uncertainty is the function of the original stoch. but also
from the stiffness matrix, and the induced bilinear form.

When $h\to 0$ the Gaussian measure collapses into a point-mass
measure at Galerkin soln.

Schematics: mesh, resolution, parameters $(x_u, \theta, h)$ $\Rightarrow$
stiffness matrix $A$ $\oplus$ $K_f$, $f$ $\Rightarrow$ solution
$u_h$.

Given data $y\in \mathbb{R}$ with (linear) measurement error from some
\textbf{true DGP}. The dgp colud be described (?) had we known
better. From FEM we get a possibly severely misspecified soln.

$$
\eta
    = \rho \Phi u_h + (\eta - \Phi u_h)
    = \rho \Phi u_h + \delta
    \,, $$
where $\eta = \Phi u^*$ for the true $u^*$ (unobserved), with
$\delta \sim \mathcal{N}(0, K)$.


Te graphical model:
$u_h$ $\Rightarrow$ observed $y$ on a grid via $\Phi$, noise $\delta$,
depending on the observation pts. on the domain $x_y$, adn \rho$.

(... dozed of)
And the mismatch between a model and reality.

$$
p(y\vert \rho, \sigma, f, h, \theta)
    = \mathcal{N}\bigl(
        \rho \Phi m^\theta_u,
        \rho^2 \Phi C_u^\theta_u \Phi m^\theta_u
        + K_f + i\sigma^2
    \bigr)
    \,, $$


Read \url{arxiv:1905.06391}. Not probabilistic numerics, since there
is no prior on the soluitions. PRob Num models nuerical error and discretization
uncertainty.


